---
title: "Workshop 2"
author: "Ean Brinker"
date: "`r Sys.Date()`"
output: html_document
---

## Workshop 2 -- Using ggplot2 for communication

```{r}
library(tidyverse)
```


#### 3.1 Labels
Making coherent graphs requires coherent labels, which we add via the labs() function.
```{r}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class)) +
  geom_smooth(se = FALSE)+
  labs(title = "Fuel efficiency generally decreases with engine size")
```
Inside labs(), you can also add subtitles and captions with identical arguments to title.
```{r}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class)) +
  geom_smooth(se = FALSE)+
  labs(title = "Fuel efficiency generally decreases with engine size",
       subtitle = "Two seaters (sports cars) are an exception because of their light weight",
       caption = "Data from fueleconomy.gov")
```
You can also edit axes and legend titles.
```{r}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class)) +
  geom_smooth(se = FALSE)+
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    color = "Car type"
  )
```

#### 3.2 Annotations
Sometimes we want to add specific text to the plot itself, rather than to our captions. This is called annotating, and we use geom_text to do it. First, we're going to filter our data to make it more manageable. There are other functions, including nudge() that let you move your labels around.
```{r}
bestinclass <- mpg %>% 
  group_by(class) %>% 
  filter(row_number(desc(hwy)) == 1) # This creates a new dataframe that takes only the first (highest) value of highway fuel efficiency for each class of car.

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class))+
  geom_text(aes(label = model), data = bestinclass)

```

#### 3.3 Scales
Changing the default scales can help readability.
```{r}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class))

# Versus:

ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class))+
  scale_x_continuous(limits = c(0,20))+
  scale_y_continuous(limits = c(0,75))+
  scale_color_manual(values= c("green","red","limegreen","cyan","magenta","yellow","blue"))
```

#### 3.4 Axis ticks
You can also change the axis scale ticks, or ignore labels altogether.
```{r}
ggplot(mpg, aes(displ, hwy))+
  geom_point()+
  scale_y_continuous(breaks = seq(15, 40, by = 5)) # This is a lazy way of writing out every number in sequences of 5 between 15 and 40.

ggplot(mpg, aes(displ, hwy))+
  geom_point()+
  scale_x_continuous(labels = NULL)+
  scale_y_continuous(labels = NULL)

```

#### 3.5 Legends and color schemes
Changing legend position is an easy addition to your code.
```{r}
base <- ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "right") #This is default
base + theme(legend.position = "bottom")
base + theme(legend.position = "none")

```

#### 3.6 Replacing a scale
Swapping continuous position scales and color scales are pretty common. We can easily plot transformations in R.
```{r}
ggplot(diamonds, aes(carat,price))+
  geom_bin2d()+
  scale_x_log10()+
  scale_y_log10()
```
Color scaling is also important.
```{r}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=drv))

# Versus:

ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=drv))+
  scale_color_brewer(palette ="Set1")

```
You can also change shapes too, to make sure your chart is distinguishable in black and white.
```{r}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=drv, shape = drv))+
  scale_color_brewer(palette ="Set1")

```
There's loads of ways to change color, and loads of packages. You can set them manually:
```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))
```
Or by stealing them from packages like viridis, which has become very popular:
```{r}
#install.packages("viridis")
#install.packages("hexbin")
library(viridis)
library(hexbin)

df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```

#### 3.7 Themes
Finally, you can customize an entire theme for your plot. ggplot2 has eight default themes, but there's loads out there, including packages like ggthemes.
```{r}
ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class))+
  geom_smooth(se = FALSE)+
  theme_bw()

ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class))+
  geom_smooth(se = FALSE)+
  theme_classic()

ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class))+
  geom_smooth(se = FALSE)+
  theme_dark()

ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class))+
  geom_smooth(se = FALSE)+
  theme_gray()

ggplot(mpg, aes(displ, hwy))+
  geom_point(aes(color=class))+
  geom_smooth(se = FALSE)+
  theme_light()


```
You can also create your own themes to standardize your plots with minimal coding. Here's one that Nick made to show the elements available for editing:
```{r}
theme(panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial")) 

```






