---
title: "Workshop 5"
author: "Ean Brinker"
date: "`r Sys.Date()`"
output: html_document
---

## Workshop 5: Spatial data in R

This workshop is designed to give you the basics of GIS and spatial data wrangling in R. This is obviously not comprehensive, but will give us a good taste.

Benefits of coding GIS, rather than using a GUI:
1. You can iteratively improve your workflow until it does exactly what you want. It's easy to remove unnecessary steps.
2. You can version control and save your work via Github.
3. You can instantly reproduce your workflow at a later date.
4. You have a easy-to-interface project to add other elements of R, such as statistical modeling.
5. R is free!

```{r}
#install.packages("sf")
#install.packages("terra")
#install.packages("tmap")

library(tidyverse)
library(sf)
library(terra)
library(tmap)

```

#### 6.6 Loading spatial data

Now it is time to load the data into R. We are going to use the readr package from tidyverse, which performs extra checks on data consistency over base R.
```{r}
library(readr)

dat <- read_csv("Data Science Module/data/data-for-course/copepods_raw.csv")

dat
```

#### 6.7 Data exploration

Knowing very little about a dataset can be problematic. Getting data ffrom an outside source means we should take some time to explore it.

##### 6.7.1 Check Coordinates

Let's check the coordinate data, make sure we can see what it looks like.
```{r}
library(ggplot2)

ggplot(dat)+
  aes(x = longitude, y = latitude, color = richness_raw)+
  geom_point()

```

This is a super simple chart, but it lacks critical map components, such as a real-world projection.

```{r}
ggplot(dat)+
  aes(x = latitude, y = richness_raw)+
  geom_point()+
  stat_smooth()

```
The data changes strangely at latitude -40. What could be going on? This is why we explore data. 

#### 6.8 Getting going with maps

For this, we will turn our point data into a spacially referenced data fram with the sf package.
```{r}
sdat <- st_as_sf(dat, coords = c("longitude", "latitude"), 
                 crs = 4326)
```

#### 6.9 Coordinate reference systems

Coordinate reference systems are the basis of what aligns our map. They compensate for the lumpy, spherical form of the Earth on a 2D plane. In this case, we are using crs4326, also known as WGS84, a very common CRS.
```{r}
crs4326 <- st_crs(4326)

crs4326 # look at the whole CRS

crs4326$Name # pull out just the name of the crs

crs4326$wkt # crs in well-known text format
```

#### 6.10 Feature Collection

Let's loot at what we have with sdat, our spatial dataframe.
```{r}
sdat
```
Fun fact, our sdat is now essentially a shapefile. Welcome back to GIS!

#### 6.11 Cartography

Nowe we get into the real mapping. The sf package has some simple plotting features we can use.
```{r}
plot(sdat["richness_raw"])
 
# This will only plot the richness_raw column. In sf, we can use square brackets to select a single column, rather than plotting everytthing at once.

plot(sdat)

```

#### 6.12 Thematic maps for communication

So far, this module has used ggplot2 for all of our plots and data viz; however, there are tons of data viz packages out there with different functionalities. Now, we're gonna use tmap. Similar to ggplot2, tmap starts with a base layer that we can add on to.
```{r}
# Using tmap:

tm1 <- tm_shape(sdat)+
  tm_dots(col = "richness_raw", palette="BuGn")

# You can change the color palette with the palette options. Use tmaptools::palette_explorer() to see your options.

# Save your map to your working directory with tmap_save.

tmap_save(tm1, filename = "Richness-map.png", 
          width = 600, height = 600)

```

#### 6.13 Mapping spatial polygons as layers

We can read shapefiles (.shp) directly into R with the st_read command.
```{r}
aus <- st_read("Data Science Module/data/data-for-course/spatial-data/Aussie/Aussie.shp")

aus

shelf <- st_read("Data Science Module/data/data-for-course/spatial-data/aus_shelf/aus_shelf.shp")

shelf

```

Mapping polygons is very easy with tmap.
```{r}
tm_shape(shelf)+
  tm_polygons()

# Let's expand the map. The argument bbox sets our boundary box of our map. In this case, we'll set it to the limit of our data.

tm_shape(shelf, bbox = sdat) + 
  tm_polygons() +
  tm_shape(aus) + 
  tm_polygons() + 
  tm_shape(sdat) + 
  tm_dots()

```

#### 6.14 Exploring tmap

Now we can try customizing our map. We can use tmap_style() to change our themes, like in ggplot. We can also use the tmap vignette to check out some tmap basics.
```{r}
vignette('tmap-getstarted')
```

Time to upgrade.
```{r}
tmap_mode("view")
tm_basemap("CartoDB.VoyagerNoLabels")+
  tm_shape(shelf, bbox = sdat) + 
  tm_polygons(alpha = 0.2, border.alpha = 1, col="skyblue4") +
  tm_shape(aus) + 
  tm_polygons(alpha = 0, border.alpha = 1) + 
  tm_shape(sdat) + 
  tm_bubbles(size="richness_raw", col="richness_raw", border.lwd = NA, palette="plasma")+
   tm_tiles("Stamen.TonerLabels")
```



