---
title: "Workshop 4"
author: "Ean Brinker"
date: "`r Sys.Date()`"
output: html_document
---

## Workshop 4 - Data Wrangling in R

This workshop focuses on preparing tabular data for use in plotting, statistic modeling, or summarizing. Today's workshop uses tidyr and the tibble object from tidyverse.


```{r}
library(tidyverse)
```

Tibbles are easily created by converting existing dataframes, as below:
```{r}
iris
str(iris) #Iris is a basic datafram included in R

as_tibble(iris)
```

Typically, we import data into R in a .csv file. To create a tibble, we're going to build it from scratch. A nice feature of tibbles is the ability to refer to variables you've just created, as below:
```{r}
tibble(
  x = 1:5,
  y = 1,
  z = x^2 + y # Even though we've not defined x and y beforehand, the tibble can recognize them as variables and make this frame happen.
)
```

How easy was that compared to the old data.frame function? The below code throws up any number of errors...
```{r}
# data.frame(c(
  # x = 1:5,
  # y = 1,
  # z = x^2 + y
# ))

```

Tibbles can also have name that normally break R and its dataframes--names that include spaces or special characters that don't work in R. We recommend still using underscores or periods in your titles rather than spaces, but it can be very useful for making comprehensible data labels. See below for examples to seperate code from labels using backticks (`). This is not an apostrophe.
```{r}
tb <- tibble(
  `:)` = "smile",
  ` ` = "space",
  `2000` = "number" 
)
tb
```

We can also use tribbles, which stands for transposed tibble (not a critter from Star Trek). These are great for doing data entry within your code.
```{r}
tribble(
  ~x, ~y, ~z,
  #--|--|--- By using the ~ symbol, we can define which column each piece of data belongs to
  "a", 2, 3.6,
  "b", 1, 8.5
)

```

Why are tibbles so much more useful? Well, tibbles tend to be much better at viewing data than a data frame.
```{r}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
)
```

Sometimes we need to see more data than the default rows or columns. We can do that with the print() function. The n argument determines how many rows you see, while width refers to the number of columns shown. Set width = Inf to see infinite (all) rows.
```{r}
#install.packages("nycflights13")
library(nycflights13)
nycflights13::flights %>% 
  print(n = 10, width = Inf)

```

if you always want certain options to be defaults, you can set global options for your R session like this:
```{r}
options(tibble.width = Inf)
```

Alright, now for more tibbling. We're just gonna call it a dataframe from now on.

```{r}
df <- tibble(
  x = runif(5),
  y = rnorm(5)
)

# Extract by name
df$x

df[["x"]]

# Extract by row position
df[[1]]

# Extract by exact position
df[[2,2]]

```

These manipulations are super important to pull out specific bits of data--maybe you need a label, or piece of text to highlight a specific value.

We can also use data pipes ( %>% ), which we'll get to later. The shortcut is Ctrl-Shift-m. We also need to include a period as a placeholder in base R. Examples:

```{r}
df %>% .$x

df %>% .[["x"]]
```

Tibbles also don't do patrial matching, so if you type in a variable wrong it will generate a warning.

```{r}
df <- tibble(
    xxx = runif(5),
    y = rnorm(5)
)

df$xx
```

Sometimes, tibbles don't play nice with older functions in R. If you need to, you can convert back to a standard data frame with as.data.frame().

Let's try figuring out this section and commenting it:
```{r}
df <- data.frame(abc = 1, xyz = "a")
df
df$x #call by name; A tibble will still find this column starting with x and return the values
df[, "xyz"] #call by exact position; this ignores the first position and returns all the XYZ values
```

#### How can I import data?

Importing data is critical in R--we're gonna do this for our QFISH assignment. We will demonstrate by building a .csv in the read_csv() function.
```{r}
read_csv("a,b,c,
1,2,3,
4,5,6")
```

In this case, read_csv() uses the first line of your data as column names. You can add a skip argument to fix this, or comment arguments.
```{r}
read_csv("The first line of metadata
  The second line of metadata
  x,y,z
  1,2,3", skip = 2)

read_csv("# A comment I want to skip
  x,y,z
  1,2,3", comment = "#")

```
 You can also tell read.csv to ignore column names.
```{r}
read_csv("1,2,3\n4,5,6", col_names = FALSE) #\n is a shortcut to make a new line, rather than hitting enter
```
 You can pass it a character vector for names instead, and tell it what values represent null or nonexistent values.
```{r}
read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z"))

read_csv("a,b,c\n1,2,.", na = ".")
```
 
 #### 5.5 Tidying data using Tidyr
 
 Data is often not tidy--see the tables below, where only table 1 is tidy. Can you see why?
```{r}
table1

table2

table3

table4a

table4b
```
 
 There are three rules for tidy data:
 1. Each variable has its own column.
 2. Each observation must have its own row
 3. Each value must how its own cell.
 
 This simplifies down to:
 1. Put each dataset in a tibble
 2. Put each variable in a column.
 
 Here are some examples of how to modify tidy data:
```{r}
# Compute rate per 10,000
table1 %>% 
  mutate(rate = cases / population * 10000)

# Compute cases per year
table1 %>% 
  count(year, wt = cases)

# Visualise changes over time
library(ggplot2)
ggplot(table1, aes(year, cases)) + 
  geom_line(aes(group = country), colour = "grey50") + 
  geom_point(aes(colour = country))

```
 
 Let's look at how to fix ugly tables. Let's start with tablee4a.
```{r}
table4a

table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")

# pivot_longer() makes datasets longer by increasing the number of rows and decreasing the number of columns

table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")

```
 
To combine them, we can use the left_join fucntion from dplyr.
```{r}
tidy4a <- table4a %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "cases")

tidy4b <- table4b %>% 
  pivot_longer(c(`1999`, `2000`), names_to = "year", values_to = "population")

left_join(tidy4a, tidy4b)
```

 Pivot_wider does the opposite, making more columns and fewer rows:
```{r}
table2

table2 %>%
    pivot_wider(names_from = type, values_from = count)

```
 
 What is the issue with table3? We've got to separate the column "rate" into two, "cases" and "population". There's a separate() function for that.
```{r}
table3

table3 %>% 
  separate(rate, into = c("cases", "population")) #Note: separate will default separate on a non_numeric character. You can set this in the arguments.

table3 %>% 
  separate(rate, into = c("cases", "population"), sep = "/")

# However, the separate() function defaults to making the columns characters, not necessarily number values. We can use the convert argument to allow it to convert them for us.

table3 %>% 
  separate(rate, into = c("cases", "population"), convert = TRUE)

```
 
We can also use the sep argument to direct separate where to divide the data. The integer value counts positions starting from the far left or uses negative values to count from the far right back. This can make our data more messy, but also can be very useful.
```{r}
table3 %>% 
  separate(year, into = c("century", "year"), sep = 2)
```
 
The inverse of separate() is unite().
```{r}
table5 %>% 
  unite(new, century, year, sep = "") # We need to add the sep argument because we don't want any separator--the default is an underscore
```

Missing data can be a big problem. Below, we have two missing values-- the 4th quarer of 2015 and the 1st quarter of 2016.
```{r}
stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
)
```
 
Right now, those values are implicit missing values. They're implied, but they're not defined as missing. Let's put years in the columns to make them explicit.
```{r}
stocks %>% 
  pivot_wider(names_from = year, values_from = return)

# Now, those missing values are a valid row of data in our tibble.

stocks %>% 
  pivot_wider(names_from = year, values_from = return) %>% 
  pivot_longer(
    cols = c(`2015`, `2016`), 
    names_to = "year", 
    values_to = "return", 
    values_drop_na = TRUE
  )

# You can also use the complete() argument to make those values explicit. In this command, R recognizes that your data is the entire set, and any missing values are meant to be missing.

stocks %>% 
  complete(year, qtr)

```

The fill() function can be used to fill in missing values that were meant to be carried forward.

```{r}
treatment <- tribble(
  ~ person,           ~ treatment, ~response,
  "Derrick Whitmore", 1,           7,
  NA,                 2,           10,
  NA,                 3,           9,
  "Katherine Burke",  1,           4
)
treatment

treatment %>% 
  fill(person)
```






